Projecting genetic associations through gene expression patterns highlights disease etiology and drug mechanisms

### 2023-06-13 output/latex/manuscript.tex
change title to supplementary information
introduction: change final paragraph according to author checklist

Here we propose PhenoPLIER, an omnigenic approach that provides a gene module perspective to genetic studies.
The flexibility of our method allows integrating different data modalities into the same representation for a joint analysis.
In this work, we show that this module perspective can infer how groups of functionally-related genes influence complex traits, detect shared and distinct transcriptomic properties among traits, and predict how pharmacological perturbations affect genes' activity to exert their effects.
PhenoPLIER maps gene-trait associations and drug-induced transcriptional responses into a common latent representation.
For this, we integrated thousands of gene-trait associations (using TWAS from PhenomeXcan\textsuperscript{\protect\hyperlink{ref-lY5ln3dB}{42}}) and transcriptional profiles of drugs (from LINCS L1000\textsuperscript{\protect\hyperlink{ref-F7lIlh2N}{43}}) into a low-dimensional space learned from public gene expression data on tens of thousands of RNA-seq samples (recount2\textsuperscript{\protect\hyperlink{ref-6SPTvFXq}{19},\protect\hyperlink{ref-14rnBunuZ}{44}}).
We used a latent representation defined by a matrix factorization approach\textsuperscript{\protect\hyperlink{ref-14rnBunuZ}{44},\protect\hyperlink{ref-Ki2ij7zE}{45}} that extracts gene modules with certain sparsity constraints and preferences for those that align with prior knowledge (pathways).
When mapping gene-trait associations to this reduced expression space, we observed that diseases were significantly associated with gene modules expressed in relevant cell types: such as hypothyroidism with T cells, corneal endothelial cells with keratometry measurements, hematological assays on specific blood cell types, plasma lipids with adipose tissue, and neuropsychiatric disorders with different brain cell types.
Moreover, since PhenoPLIER can use models derived from large and heterogeneous RNA-seq datasets, we could also identify modules associated with cell types under specific stimuli or disease states.
We observed that significant module-trait associations in PhenomeXcan (our discovery cohort) replicated in the Electronic Medical Records and Genomics (eMERGE) network phase III\textsuperscript{\protect\hyperlink{ref-gZAOkumx}{27},\protect\hyperlink{ref-wfqjCerX}{46}} (our replication cohort).
Furthermore, we performed a CRISPR screen to analyze lipid regulation in HepG2 cells.
We observed more robust trait associations with modules than with individual genes, even when single genes known to be involved in lipid metabolism did not reach genome-wide significance.
Compared to a single-gene approach, our module-based method also better predicted FDA-approved drug-disease links by capturing tissue-specific pathophysiological mechanisms linked with the mechanism of action of drugs (e.g., niacin with cardiovascular traits via a known immune mechanism).
This improved drug-disease prediction suggested that modules may provide a better means to examine drug-disease relationships than individual genes.
Finally, exploring the phenotype-module space revealed stable trait clusters associated with relevant tissues, including a complex branch involving lipids with cardiovascular, autoimmune, and neuropsychiatric disorders.
In summary, instead of considering single genes associated with different complex traits, PhenoPLIER incorporates groups of genes that act together to carry out different functions in specific cell types.
This approach improves robustness in detecting and interpreting genetic associations, and here we show how it can prioritize alternative and potentially more promising candidate targets even when known single gene associations are not detected.
The approach represents a conceptual shift in the interpretation of genetic studies.
It has the potential to extract mechanistic insight from statistical associations to enhance the understanding of complex diseases and their therapeutic modalities.

@@

\hypertarget{supplementary-material}{%
\subsection{Supplementary material}\label{supplementary-material}}

>>>>

Here we propose PhenoPLIER, an omnigenic approach that provides a gene module perspective to genetic studies.
The flexibility of our method allows integrating different data modalities into the same representation for a joint analysis.
We show that this module perspective can infer how groups of functionally-related genes influence complex traits, detect shared and distinct transcriptomic properties among traits, and predict how pharmacological perturbations affect genes' activity to exert their effects.
PhenoPLIER maps gene-trait associations and drug-induced transcriptional responses into a common latent representation.
For this, we integrate thousands of gene-trait associations (using TWAS from PhenomeXcan\textsuperscript{\protect\hyperlink{ref-lY5ln3dB}{42}}) and transcriptional profiles of drugs (from LINCS L1000\textsuperscript{\protect\hyperlink{ref-F7lIlh2N}{43}}) into a low-dimensional space learned from public gene expression data on tens of thousands of RNA-seq samples (recount2\textsuperscript{\protect\hyperlink{ref-6SPTvFXq}{19},\protect\hyperlink{ref-14rnBunuZ}{44}}).
We use a latent representation defined by a matrix factorization approach\textsuperscript{\protect\hyperlink{ref-14rnBunuZ}{44},\protect\hyperlink{ref-Ki2ij7zE}{45}} that extracts gene modules with certain sparsity constraints and preferences for those that align with prior knowledge (pathways).
When mapping gene-trait associations to this reduced expression space, we observe that diseases are significantly associated with gene modules expressed in relevant cell types: such as hypothyroidism with T cells, corneal endothelial cells with keratometry measurements, hematological assays on specific blood cell types, plasma lipids with adipose tissue, and neuropsychiatric disorders with different brain cell types.
Moreover, since PhenoPLIER can use models derived from large and heterogeneous RNA-seq datasets, we can also identify modules associated with cell types under specific stimuli or disease states.
We observe that significant module-trait associations in PhenomeXcan (our discovery cohort) replicated in the Electronic Medical Records and Genomics (eMERGE) network phase III\textsuperscript{\protect\hyperlink{ref-gZAOkumx}{27},\protect\hyperlink{ref-wfqjCerX}{46}} (our replication cohort).
Furthermore, we perform a CRISPR screen to analyze lipid regulation in HepG2 cells.
We observe more robust trait associations with modules than with individual genes, even when single genes known to be involved in lipid metabolism did not reach genome-wide significance.
Compared to a single-gene approach, our module-based method also better predicts FDA-approved drug-disease links by capturing tissue-specific pathophysiological mechanisms linked with the mechanism of action of drugs (e.g., niacin with cardiovascular traits via a known immune mechanism).
This improved drug-disease prediction suggests that modules may provide a better means to examine drug-disease relationships than individual genes.
Finally, exploring the phenotype-module space reveals stable trait clusters associated with relevant tissues, including a complex branch involving lipids with cardiovascular, autoimmune, and neuropsychiatric disorders.
In summary, instead of considering single genes associated with different complex traits, PhenoPLIER incorporates groups of genes that act together to carry out different functions in specific cell types.
This approach improves robustness in detecting and interpreting genetic associations, and here we show how it can prioritize alternative and potentially more promising candidate targets even when known single gene associations are not detected.
The approach represents a conceptual shift in the interpretation of genetic studies.
It has the potential to extract mechanistic insight from statistical associations to enhance the understanding of complex diseases and their therapeutic modalities.

@@

\hypertarget{supplementary-information}{%
\subsection{Supplementary information}\label{supplementary-information}}
